---
title: "Multilevel Models: Interest Group Prominence in Congressional Speech"
author: "Kaleb Mazurek"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
always_allow_html: true
output:
  github_document:
    toc: true
    html_preview: false
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```


---

# Overview

This R Markdown document implements the primary inferential models from my thesis on interest group prominence in congressional floor speeches. Three sets of generalized linear mixed-effects models (GLMMs) test what predicts whether an interest group mention is prominent (cited as authoritative) versus routine.

**Models:**

- **Model A**: Issue salience (does policy salience predict prominence?)
- **Model B**: Politician characteristics (seniority, party, chamber, legislative activity)
- **Model C**: Organizational resources (age, lobbying expenditure, policy scope)

Random effects: Organization ID and Policy Area (crossed).

Data: 19,165 mentions of 500+ advocacy organizations in the 114th-115th Congress. Prominence classified by SVM at ~81% accuracy.

**Key findings:**

1. Medium-salience policy areas predict prominence better than high-salience areas
2. Senior legislators are less likely to cite groups prominently
3. External lobbyists modestly increase prominence, but the effect is not large

For the full thesis manuscript, see `../legacy/5. Visualization and Reporting/Thesis_UvA_Kaleb_Mazurek.pdf`.

---

# Setup and Data Loading

```{r libraries}
# Core data manipulation
library(tidyverse)
library(haven)

# Statistical modeling
library(lme4)        # Mixed-effects models
library(lmerTest)    # p-values for mixed models
library(broom.mixed) # Tidy model outputs

# Model comparison and diagnostics
library(performance)
library(see)
library(car)

# Visualization
library(ggplot2)
library(ggeffects)
library(sjPlot)
library(patchwork)

# Tables
library(knitr)
library(kableExtra)
library(gt)

# Set theme
theme_set(theme_minimal(base_size = 12))

# Display options
options(
  digits = 4,
  scipen = 999,
  knitr.kable.NA = ''
)
```

```{r config}
# Configuration
CONFIG <- list(
  # File paths (relative to analysis/ directory)
  DATA_DIR = file.path("..", "data", "multi_level_data"),
  OUTPUT_DIR = file.path("..", "output", "figures"),

  # Data file
  LEVEL1_FILE = "level1_FINAL.csv",

  # Organization to exclude (if applicable)
  EXCLUDED_ORG_ID = NULL,

  # Random seed
  SEED = 42,

  # Reference categories for modeling
  REF_CHAMBER = "House of Representatives",
  REF_PARTY = "Democrat",
  REF_SALIENCY = "low",
  REF_ORGTYPE = "Business Interests",
  REF_MSHIP = "Association of Institutions",
  REF_TERM = "First Year"
)

# Create output directory
dir.create(CONFIG$OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)

# Set seed for reproducibility
set.seed(CONFIG$SEED)
```

```{r data-loading-functions}
#' Load and perform initial filtering of level1 dataset
#'
#' @param filepath Path to CSV file
#' @param excluded_org_id Organization ID to exclude (optional)
#' @return Filtered dataframe
load_data <- function(filepath, excluded_org_id = NULL) {
  # Only read columns used by the models to reduce memory (~175 cols -> ~15)
  needed_cols <- c(
    "level1_prominence", "level1_org_id", "level1_issue_area",
    "level1_chamber_x", "level1_partyHistory", "level1_ABBREVCAT",
    "level1_MSHIP_STATUS11", "level1_seniority", "level1_bills_sponsored",
    "level1_issue_maximal_overlap", "level1_YEARS_EXISTED", "level1_OUTSIDE11",
    "level1_congress",
    # Source columns for derived variables (term_status, saliency_category)
    "level1_year_week", "level1_termBeginYear", "level1_termEndYear",
    "level1_issue_area_salience"
  )
  df <- read_csv(filepath, col_select = all_of(needed_cols), show_col_types = FALSE)

  if (!is.null(excluded_org_id)) {
    df <- df %>% filter(level1_org_id != excluded_org_id)
  }

  message(sprintf("Loaded %s rows, %d columns", format(nrow(df), big.mark = ","), ncol(df)))
  return(df)
}

#' Recode organization category variable
#'
#' Collapses detailed categories into broader groupings:
#' - Business Interests
#' - Government Interests
#' - Non-Business Interests
recode_abbrevcat <- function(df) {
  # Original to readable name mapping
  name_mapping <- c(
    "(1) Corporations" = "Corporations",
    "(13) Social welfare or poor" = "Social Welfare or Poor",
    "(14) State and local governments" = "State and Local Governments",
    "(16) Other" = "Other",
    "(2) Trade and other business associations" = "Trade and Business Associations",
    "(3) Occupational associations" = "Occupational Associations",
    "(4) Unions" = "Unions",
    "(5) Education" = "Education",
    "(6) Health" = "Health",
    "(7) Public interest" = "Public Interest",
    "(8) Identity groups" = "Identity Groups"
  )

  df <- df %>%
    mutate(
      level1_ABBREVCAT = name_mapping[level1_ABBREVCAT],
      # Remove Corporations
      level1_ABBREVCAT = if_else(level1_ABBREVCAT == "Corporations", NA_character_, level1_ABBREVCAT)
    ) %>%
    filter(!is.na(level1_ABBREVCAT)) %>%
    mutate(
      # Collapse into broader categories
      level1_ABBREVCAT = case_when(
        level1_ABBREVCAT %in% c("Trade and Business Associations", "Corporations") ~ "Business-Oriented Interests",
        level1_ABBREVCAT == "State and Local Governments" ~ "Government Interests",
        TRUE ~ "Non-business/nongovernment"
      ),
      # Final collapse
      level1_ABBREVCAT = case_when(
        level1_ABBREVCAT == "Business-Oriented Interests" ~ "Business Interests",
        level1_ABBREVCAT == "Government Interests" ~ "Government Interests",
        TRUE ~ "Non-Business Interests"
      ),
      # Binary indicator
      business_interest = as.integer(level1_ABBREVCAT == "Business Interests")
    )

  message("ABBREVCAT distribution:")
  print(table(df$level1_ABBREVCAT))

  return(df)
}

#' Recode membership status variable
#'
#' Collapses categories into:
#' - Association of Individuals
#' - Association of Institutions
#' - Other
recode_membership_status <- function(df) {
  name_mapping <- c(
    "(1) Institution" = "Institution",
    "(2) Association of individuals" = "Association of Individuals",
    "(3) Association of institutions" = "Association of Institutions",
    "(4) Government or association of governments" = "Government or Association of Governments",
    "(5) Mixed" = "Mixed",
    "(6) Other" = "Other",
    "(9) Cant tell or DK" = "Can't Tell"
  )

  df <- df %>%
    mutate(
      level1_MSHIP_STATUS11 = name_mapping[level1_MSHIP_STATUS11],
      level1_MSHIP_STATUS11 = case_when(
        level1_MSHIP_STATUS11 == "Association of Individuals" ~ "Association of Individuals",
        level1_MSHIP_STATUS11 %in% c("Institution", "Association of Institutions",
                                      "Government or Association of Governments") ~ "Association of Institutions",
        TRUE ~ "Other"
      )
    )

  message("Membership status distribution:")
  print(table(df$level1_MSHIP_STATUS11))

  return(df)
}

#' Create term status variable
#'
#' Categories:
#' - First Year: Mention in first year of term
#' - Year Before Term End: Mention in year before term ends
#' - Other: All other years
create_term_status <- function(df) {
  df <- df %>%
    mutate(
      mention_year = as.integer(substr(as.character(level1_year_week), 1, 4)),
      year_before_termEnd = as.integer(!is.na(mention_year) & mention_year == (level1_termEndYear - 1)),
      first_year_term = as.integer(!is.na(mention_year) & mention_year == level1_termBeginYear),
      term_status = case_when(
        is.na(first_year_term) | is.na(year_before_termEnd) ~ NA_character_,
        first_year_term == 1 & year_before_termEnd == 0 ~ "First Year",
        first_year_term == 0 & year_before_termEnd == 1 ~ "Year Before Term End",
        TRUE ~ "Other"
      )
    )

  message("Term status distribution:")
  print(table(df$term_status, useNA = "ifany"))

  return(df)
}

#' Compute issue area features for each organization
#'
#' Creates:
#' - most_common_issue_area: Mode of issue areas per org
#' - unique_issue_areas: Count of distinct issue areas per org
#' - issue_area_overlap: Whether mention is in org's most common area
compute_issue_area_features <- function(df) {
  # Most common issue area per organization
  most_common <- df %>%
    group_by(level1_org_id) %>%
    summarize(
      most_common_issue_area = {
        tbl <- table(level1_issue_area)
        if (length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
      },
      .groups = "drop"
    )

  # Unique issue areas per organization
  unique_areas <- df %>%
    group_by(level1_org_id) %>%
    summarize(unique_issue_areas = n_distinct(level1_issue_area), .groups = "drop")

  # Merge back
  df <- df %>%
    left_join(most_common, by = "level1_org_id") %>%
    left_join(unique_areas, by = "level1_org_id") %>%
    mutate(issue_area_overlap = as.integer(level1_issue_area == most_common_issue_area))

  message(sprintf("Mean unique issue areas per org: %.2f", mean(df$unique_issue_areas, na.rm = TRUE)))

  return(df)
}

#' Compute saliency measure based on issue area
#'
#' Categories: low (1-7), medium (8-14), high (15-21)
compute_saliency_measure <- function(df) {
  # Use the existing issue_area_salience column from the dataset
  df <- df %>%
    mutate(
      saliency_measure = as.numeric(level1_issue_area_salience),
      saliency_category = case_when(
        is.na(saliency_measure) ~ NA_character_,
        saliency_measure <= 7 ~ "low",
        saliency_measure <= 14 ~ "medium",
        TRUE ~ "high"
      ),
      saliency_category = factor(saliency_category, levels = c("low", "medium", "high"))
    )

  message("Saliency distribution:")
  print(table(df$saliency_category, useNA = "ifany"))

  return(df)
}

#' Clean and prepare categorical variables for modeling
clean_categorical_variables <- function(df) {
  df <- df %>%
    mutate(
      # Replace empty strings with NA
      across(c(level1_chamber_x, level1_partyHistory), ~na_if(., "")),
      # Convert to factors with explicit levels
      level1_chamber_x = factor(level1_chamber_x),
      level1_partyHistory = factor(level1_partyHistory),
      saliency_category = factor(saliency_category, levels = c("low", "medium", "high")),
      level1_issue_area = factor(level1_issue_area),
      level1_ABBREVCAT = factor(level1_ABBREVCAT),
      level1_MSHIP_STATUS11 = factor(level1_MSHIP_STATUS11),
      term_status = factor(term_status),
      # Ensure prominence is numeric (for glmer)
      level1_prominence = as.numeric(level1_prominence)
    )

  return(df)
}
```

```{r preprocessing-pipeline}
#' Complete data preprocessing pipeline
#'
#' Chains all preprocessing steps together
preprocess_data <- function(filepath, excluded_org_id = NULL) {
  message("=" %>% str_dup(60))
  message("DATA PREPROCESSING PIPELINE")
  message("=" %>% str_dup(60))

  message("\n[1/7] Loading data...")
  df <- load_data(filepath, excluded_org_id)

  message("\n[2/7] Recoding organization categories...")
  df <- recode_abbrevcat(df)

  message("\n[3/7] Recoding membership status...")
  df <- recode_membership_status(df)

  message("\n[4/7] Creating term status...")
  df <- create_term_status(df)

  message("\n[5/7] Computing issue area features...")
  df <- compute_issue_area_features(df)

  message("\n[6/7] Computing saliency measure...")
  df <- compute_saliency_measure(df)

  message("\n[7/7] Cleaning categorical variables...")
  df <- clean_categorical_variables(df)

  message("\n" %+% ("=" %>% str_dup(60)))
  message(sprintf("Preprocessing complete. Final dataset: %s rows, %d columns",
                  format(nrow(df), big.mark = ","), ncol(df)))
  message("=" %>% str_dup(60))

  return(df)
}
```

## Load and Preprocess Data

```{r load-data}
# Load actual thesis data
level1 <- preprocess_data(
  file.path(CONFIG$DATA_DIR, CONFIG$LEVEL1_FILE),
  excluded_org_id = CONFIG$EXCLUDED_ORG_ID
)

message(sprintf("Loaded dataset with %s observations", format(nrow(level1), big.mark = ",")))
```

## Data Overview

```{r data-summary}
# Summary statistics
cat("Dataset dimensions:", nrow(level1), "rows x", ncol(level1), "columns\n\n")

cat("Prominence distribution:\n")
table(level1$level1_prominence) %>% prop.table() %>% round(3) %>% print()

cat("\nSaliency category distribution:\n")
table(level1$saliency_category, useNA = "ifany") %>% print()

cat("\nOrganization type distribution:\n")
table(level1$level1_ABBREVCAT, useNA = "ifany") %>% print()

cat("\nChamber distribution:\n")
table(level1$level1_chamber_x, useNA = "ifany") %>% print()

cat("\nParty distribution:\n")
table(level1$level1_partyHistory, useNA = "ifany") %>% print()
```

---

# Model A: Issue Salience

Are interest groups mentioned in high-salience policy areas more likely to be cited as authoritative voices? Salience is measured via Google Trends data aggregated by CAP policy domain into low, medium, and high terciles. The thesis hypothesized a linear relationship (higher salience = more prominence), but the results suggest something more complicated.

## Model A: Empty Model

```{r model-empty}
# Fit empty model once (identical formula reused for all three model sets)
empty_model <- glmer(
  level1_prominence ~ 1 + (1 | level1_org_id) + (1 | level1_issue_area),
  data = level1,
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(empty_model)
```

## Model A1: Saliency Category Only

```{r model-a1}
# Set reference level
level1 <- level1 %>%
  mutate(saliency_category = relevel(factor(saliency_category), ref = "low"))

# Fit model with saliency only
model_a1 <- glmer(
  level1_prominence ~ saliency_category + (1 | level1_org_id) + (1 | level1_issue_area),
  data = level1,
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_a1)

# Display odds ratios
exp(fixef(model_a1)) %>%
  round(4) %>%
  kable(col.names = "Odds Ratio", caption = "Model A1: Saliency Effects (Odds Ratios)") %>%
  kable_styling()
```

## Model A2: Full Model with Controls

```{r model-a2}
# Set reference levels for all categorical variables
level1 <- level1 %>%
  mutate(
    level1_chamber_x = relevel(factor(level1_chamber_x), ref = CONFIG$REF_CHAMBER),
    level1_partyHistory = relevel(factor(level1_partyHistory), ref = CONFIG$REF_PARTY),
    level1_ABBREVCAT = relevel(factor(level1_ABBREVCAT), ref = CONFIG$REF_ORGTYPE),
    level1_MSHIP_STATUS11 = relevel(factor(level1_MSHIP_STATUS11), ref = CONFIG$REF_MSHIP)
  )

# Fit full model
model_a2 <- glmer(
  level1_prominence ~
    saliency_category +
    level1_chamber_x +
    level1_partyHistory +
    level1_MSHIP_STATUS11 +
    level1_ABBREVCAT +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(saliency_category, level1_chamber_x, level1_partyHistory,
                             level1_MSHIP_STATUS11, level1_ABBREVCAT),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_a2)

# Odds ratios table
or_table_a2 <- broom.mixed::tidy(model_a2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  )

or_table_a2 %>%
  kable(
    digits = 4,
    col.names = c("Term", "Odds Ratio", "95% CI Lower", "95% CI Upper", "p-value", "Sig."),
    caption = "Model A2: Full Model with Controls (Odds Ratios)"
  ) %>%
  kable_styling()
```

## Model A: Comparison

```{r model-a-comparison}
# Model fit statistics
compare_models_a <- tibble(
  Model = c("Empty", "A1: Saliency", "A2: Full"),
  AIC = c(AIC(empty_model), AIC(model_a1), AIC(model_a2)),
  BIC = c(BIC(empty_model), BIC(model_a1), BIC(model_a2)),
  LogLik = c(logLik(empty_model)[1], logLik(model_a1)[1], logLik(model_a2)[1]),
  N = c(nobs(empty_model), nobs(model_a1), nobs(model_a2))
)

compare_models_a %>%
  kable(digits = 2, caption = "Model A Comparison") %>%
  kable_styling()
```

## Visualization: Model A Results

```{r viz-model-a, fig.height=8}
# Forest plot of odds ratios
plot_model(model_a2,
           type = "est",
           show.values = TRUE,
           value.offset = 0.3,
           vline.color = "red",
           title = "Model A: Issue Salience Effects on Prominence",
           axis.title = "Odds Ratio (95% CI)") +
  theme_minimal(base_size = 12)
```

```{r viz-predicted-probs-a}
# Predicted probabilities by saliency
pred_a <- ggpredict(model_a2, terms = "saliency_category")

plot(pred_a, show_data = FALSE) +
  labs(
    title = "Predicted Probability of Prominence by Issue Salience",
    x = "Saliency Category",
    y = "Predicted Probability of Prominence"
  ) +
  theme_minimal(base_size = 12)
```

```{r cleanup-model-a, include=FALSE}
rm(model_a1, model_a2, or_table_a2, pred_a); gc()
```

## Interpretation: Model A

Medium-salience policy areas predict prominence better than high-salience areas, which was unexpected. One possibility is that high-salience issues attract so many competing voices that individual group visibility gets diluted. In medium-salience domains, the field is less crowded and individual organizations can more easily become the go-to reference. This is consistent with the concentration patterns from the exploratory analysis.

The committee specialization story is worth noting: members on specialized committees may cite groups more prominently in their jurisdiction areas, which tend to be medium-salience (agriculture, transportation, energy). High-salience issues like healthcare and immigration cut across committee boundaries, making any single group citation less strategically useful.

---

## Model B: Politician Characteristics

Do politician-level factors (seniority, party, chamber, legislative activity) predict whether they cite interest groups as authoritative voices? Model B tests variables from the politician side of the prominence dynamic, controlling for the same organizational and institutional factors as Model A.

## Model B: Empty Model

```{r model-b-empty}
# Reuse empty model from Model A (identical formula and data)
summary(empty_model)
```

## Model B1: Politician Characteristics

```{r model-b1}
# Set reference level for term status
level1 <- level1 %>%
  mutate(term_status = relevel(factor(term_status), ref = "First Year"))

model_b1 <- glmer(
  level1_prominence ~
    level1_issue_maximal_overlap +
    term_status +
    level1_bills_sponsored +
    level1_seniority +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_issue_maximal_overlap, term_status,
                             level1_bills_sponsored, level1_seniority),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_b1)

# Odds ratios
broom.mixed::tidy(model_b1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  kable(digits = 4, caption = "Model B1: Politician Characteristics (Odds Ratios)") %>%
  kable_styling()
```

## Model B2: Full Model with Controls

```{r model-b2}
model_b2 <- glmer(
  level1_prominence ~
    level1_issue_maximal_overlap +
    term_status +
    level1_bills_sponsored +
    level1_seniority +
    level1_chamber_x +
    level1_partyHistory +
    level1_MSHIP_STATUS11 +
    level1_ABBREVCAT +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_issue_maximal_overlap, term_status, level1_bills_sponsored,
                             level1_seniority, level1_chamber_x, level1_partyHistory,
                             level1_MSHIP_STATUS11, level1_ABBREVCAT),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_b2)

# Odds ratios table
broom.mixed::tidy(model_b2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Odds Ratio", "95% CI Lower", "95% CI Upper", "p-value", "Sig."),
    caption = "Model B2: Full Model (Odds Ratios)"
  ) %>%
  kable_styling()
```

## Model B: Comparison

```{r model-b-comparison}
compare_models_b <- tibble(
  Model = c("Empty", "B1: Politician Chars", "B2: Full"),
  AIC = c(AIC(empty_model), AIC(model_b1), AIC(model_b2)),
  BIC = c(BIC(empty_model), BIC(model_b1), BIC(model_b2)),
  LogLik = c(logLik(empty_model)[1], logLik(model_b1)[1], logLik(model_b2)[1]),
  N = c(nobs(empty_model), nobs(model_b1), nobs(model_b2))
)

compare_models_b %>%
  kable(digits = 2, caption = "Model B Comparison") %>%
  kable_styling()
```

## Visualization: Model B Results

```{r viz-model-b, fig.height=8}
plot_model(model_b2,
           type = "est",
           show.values = TRUE,
           value.offset = 0.3,
           vline.color = "red",
           title = "Model B: Politician-Group Linkage Effects on Prominence",
           axis.title = "Odds Ratio (95% CI)") +
  theme_minimal(base_size = 12)
```

```{r viz-seniority-effect}
# Predicted probabilities by seniority
pred_b_seniority <- ggpredict(model_b2, terms = "level1_seniority [all]")

plot(pred_b_seniority, show_data = FALSE) +
  labs(
    title = "Predicted Probability of Prominence by Member Seniority",
    x = "Years in Congress",
    y = "Predicted Probability of Prominence"
  ) +
  theme_minimal(base_size = 12)
```

```{r cleanup-model-b, include=FALSE}
rm(model_b1, model_b2, pred_b_seniority); gc()
```

## Interpretation: Model B

The seniority finding ran counter to expectations. More senior legislators are slightly less likely to afford prominence to advocacy groups, contradicting the intuition that longer-serving members would have more established group relationships. One explanation: senior members have enough personal credibility that they don't need external validators. They chair committees and shape policy directly rather than through public citations.

The null findings are informative. Electoral cycle timing, bill sponsorship, and issue overlap do not clearly predict prominence-granting behavior. This suggests that the politician side of the prominence dynamic is less systematic than theoretical models assume, or that the relevant politician-level factors haven't been identified yet. The weak issue overlap effect suggests members do cite groups somewhat more in their areas of expertise, but specialization alone doesn't explain the pattern.

---

## Model C: Organizational Resources

Do organizational characteristics (age, lobbying expenditure, policy scope, use of external lobbyists) predict prominence? Model C tests whether well-resourced organizations dominate legislative discourse, as classic pluralist theory predicts.

## Model C: Empty Model

```{r model-c-empty}
# Reuse empty model from Model A (identical formula and data)
summary(empty_model)
```

## Model C1: Organizational Characteristics

```{r model-c1}
model_c1 <- glmer(
  level1_prominence ~
    level1_YEARS_EXISTED +
    level1_OUTSIDE11 +
    unique_issue_areas +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_YEARS_EXISTED, level1_OUTSIDE11, unique_issue_areas),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_c1)

# Odds ratios
broom.mixed::tidy(model_c1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  kable(digits = 4, caption = "Model C1: Organizational Characteristics (Odds Ratios)") %>%
  kable_styling()
```

## Model C2: Full Model with Controls

```{r model-c2}
model_c2 <- glmer(
  level1_prominence ~
    level1_YEARS_EXISTED +
    level1_OUTSIDE11 +
    unique_issue_areas +
    level1_chamber_x +
    level1_partyHistory +
    level1_MSHIP_STATUS11 +
    level1_ABBREVCAT +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_YEARS_EXISTED, level1_OUTSIDE11, unique_issue_areas,
                             level1_chamber_x, level1_partyHistory,
                             level1_MSHIP_STATUS11, level1_ABBREVCAT),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_c2)

# Odds ratios table
broom.mixed::tidy(model_c2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Odds Ratio", "95% CI Lower", "95% CI Upper", "p-value", "Sig."),
    caption = "Model C2: Full Model (Odds Ratios)"
  ) %>%
  kable_styling()
```

## Model C: Comparison

```{r model-c-comparison}
compare_models_c <- tibble(
  Model = c("Empty", "C1: Org Characteristics", "C2: Full"),
  AIC = c(AIC(empty_model), AIC(model_c1), AIC(model_c2)),
  BIC = c(BIC(empty_model), BIC(model_c1), BIC(model_c2)),
  LogLik = c(logLik(empty_model)[1], logLik(model_c1)[1], logLik(model_c2)[1]),
  N = c(nobs(empty_model), nobs(model_c1), nobs(model_c2))
)

compare_models_c %>%
  kable(digits = 2, caption = "Model C Comparison") %>%
  kable_styling()
```

## Visualization: Model C Results

```{r viz-model-c, fig.height=8}
plot_model(model_c2,
           type = "est",
           show.values = TRUE,
           value.offset = 0.3,
           vline.color = "red",
           title = "Model C: Organizational Characteristics Effects on Prominence",
           axis.title = "Odds Ratio (95% CI)") +
  theme_minimal(base_size = 12)
```

```{r viz-lobbyist-effect}
# Predicted probabilities by external lobbyist use
pred_c_lobbyist <- ggpredict(model_c2, terms = "level1_OUTSIDE11")

plot(pred_c_lobbyist, show_data = FALSE) +
  labs(
    title = "Predicted Probability of Prominence by External Lobbyist Use",
    x = "Uses External Lobbyists",
    y = "Predicted Probability of Prominence"
  ) +
  scale_x_continuous(breaks = c(0, 1), labels = c("No", "Yes")) +
  theme_minimal(base_size = 12)
```

```{r cleanup-model-c, include=FALSE}
rm(model_c1, model_c2, pred_c_lobbyist, empty_model); gc()
```

## Interpretation: Model C

The external lobbyist finding was the most surprising result. I hypothesized that reliance on external lobbyists would not significantly increase prominence, following Grossmann's (2012) argument that outsourcing advocacy signals weak internal capacity. The opposite held: organizations employing external lobbyists had significantly higher prominence odds (p = 0.001). Professional intermediaries appear to play a meaningful role in positioning organizations within legislative discourse.

Organizational age showed no significant effect, which challenges the assumption that longevity translates to being treated as a prominent voice. Whatever drives prominence, it is not simply a matter of having been around for a long time. Policy breadth showed a positive but non-significant trend, suggesting that a broader agenda creates more intersections with legislative debate but doesn't guarantee prominent treatment.

The overall picture from all three model sets: prominence is driven more by organizational positioning and the structure of the policy environment than by the characteristics of individual politicians.

---

# Overall Results Summary

## Cross-Model Comparison

```{r overall-comparison}
# Combine all model comparisons
all_comparisons <- bind_rows(
  compare_models_a %>% mutate(Model_Set = "A: Salience"),
  compare_models_b %>% mutate(Model_Set = "B: Politician"),
  compare_models_c %>% mutate(Model_Set = "C: Organization")
)

all_comparisons %>%
  select(Model_Set, Model, AIC, BIC, LogLik, N) %>%
  kable(digits = 2, caption = "Overall Model Comparison") %>%
  kable_styling() %>%
  pack_rows("Model Set A: Issue Salience", 1, 3) %>%
  pack_rows("Model Set B: Politician Characteristics", 4, 6) %>%
  pack_rows("Model Set C: Organizational Resources", 7, 9)
```

## Visualization: Model Fit Comparison

```{r viz-model-comparison, fig.width=12, fig.height=6}
# AIC comparison
p1 <- ggplot(all_comparisons, aes(x = Model, y = AIC, color = Model_Set, group = Model_Set)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "AIC Comparison (Lower is Better)", y = "AIC", color = "Model Set") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# BIC comparison
p2 <- ggplot(all_comparisons, aes(x = Model, y = BIC, color = Model_Set, group = Model_Set)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "BIC Comparison (Lower is Better)", y = "BIC", color = "Model Set") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots
p1 + p2 + plot_layout(guides = "collect")
```

---


---

# Discussion

The three model sets converge on a finding that prominence operates through different channels than traditional lobbying access:

- **Salience paradox**: Groups gain prominence in medium-salience areas, not high-salience ones. Legislators may avoid citing external validators on polarized issues where group names become partisan signals.
- **Seniority reversal**: Junior members cite groups more prominently, possibly because they need external validators to establish credibility.
- **Moderate resource effects**: External lobbyists increase prominence odds, but organizational age has minimal effect. Newer groups can break through.

These patterns distinguish prominence from lobbying access and policy influence as a separate dimension of interest group success (extending Grossmann, 2012; Halpin & Fraussen, 2017).

**Hypothesis mapping**: Model A partially supports H1: medium salience predicts prominence, but high salience reduces it rather than increasing it. Model B partially supports H2: seniority is significant but in the opposite direction from what was predicted. Model C rejects the null H3c: external lobbyists do significantly predict prominence (p = 0.001), contradicting the hypothesis that they would have no effect. Full hypothesis statements and theoretical grounding are in the thesis document.

## Limitations

- Congressional Record captures floor speeches only, not committee hearings or private communication
- Time window (2015-2019) predates recent polarization shifts
- Classifier accuracy (~81%) introduces measurement noise
- Cross-sectional design limits causal claims

---

# Appendix: Technical Notes

## Session Information

```{r session-info}
sessionInfo()
```

## Model Diagnostics

```{r diagnostics-a2, eval=FALSE}
# Model A2 diagnostics
check_model(model_a2)
```

```{r diagnostics-b2, eval=FALSE}
# Model B2 diagnostics
check_model(model_b2)
```

```{r diagnostics-c2, eval=FALSE}
# Model C2 diagnostics
check_model(model_c2)
```

## Export Results

```{r export-results}
# Export model comparison table (individual models already freed to save memory)
all_comparisons %>%
  write_csv(file.path(CONFIG$OUTPUT_DIR, "model_comparison.csv"))

message("Results exported to ", CONFIG$OUTPUT_DIR)
```

---

# References

Arnold, R. D. (1990). *The Logic of Congressional Action*. Yale University Press.

Fenno, R. F. (1978). *Home Style: House Members in Their Districts*. Little, Brown.

Grossmann, M. (2012). *The Not-So-Special Interests: Interest Groups, Public Representation, and American Governance*. Stanford University Press.

Halpin, D., & Fraussen, B. (2017). Conceptualizing the policy engagement of interest groups: Involvement, access and prominence. *European Journal of Political Research*, 56(3), 467-483.

Mayhew, D. R. (1974). *Congress: The Electoral Connection*. Yale University Press.

---

**Document compiled**: `r format(Sys.time(), '%B %d, %Y at %I:%M %p')`
