---
title: "Whose Voice Counts? Understanding Advocacy Group Prominence in Congressional Discourse"
author: "Kaleb Mazurek"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
always_allow_html: true
output:
  github_document:
    toc: true
    html_preview: false
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```


---

# Overview

This R Markdown document implements the primary inferential models from my thesis on interest group prominence in congressional floor speeches. Three sets of generalized linear mixed-effects models (GLMMs) test what predicts whether an interest group mention is prominent (cited as authoritative) versus routine.

**Models:**

- **Model A**: Issue salience (does policy salience predict prominence?)
- **Model B**: Politician characteristics (seniority, party, chamber, legislative activity)
- **Model C**: Organizational resources (age, lobbying expenditure, policy scope)

Random effects: Organization ID and Policy Area (crossed).

Data: 19,165 mentions of 500+ advocacy organizations in the 114th-115th Congress. Prominence classified by SVM at ~81% accuracy.

**Key findings:**

1. Medium-salience policy areas predict prominence better than high-salience areas
2. Senior legislators are less likely to cite groups prominently
3. External lobbyists modestly increase prominence, but the effect is not large

For the full thesis manuscript, see `../legacy/5. Visualization and Reporting/Thesis_UvA_Kaleb_Mazurek.pdf`.

---

# Setup and Data Loading

```{r libraries}
# Core data manipulation
library(tidyverse)
library(haven)

# Statistical modeling
library(lme4)        # Mixed-effects models
library(lmerTest)    # p-values for mixed models
library(broom.mixed) # Tidy model outputs

# Model comparison and diagnostics
library(performance)
library(see)
library(car)

# Visualization
library(ggplot2)
library(ggeffects)
library(sjPlot)
library(patchwork)

# Tables
library(knitr)
library(kableExtra)
library(gt)

# Set theme
theme_set(theme_minimal(base_size = 12))

# Display options
options(
  digits = 4,
  scipen = 999,
  knitr.kable.NA = ''
)
```

```{r config}
# Configuration
CONFIG <- list(
  # File paths (relative to analysis/ directory)
  DATA_DIR = file.path("..", "data", "multi_level_data"),
  OUTPUT_DIR = file.path("..", "output", "figures"),

  # Data file
  LEVEL1_FILE = "level1_FINAL.csv",

  # Organization to exclude (if applicable)
  EXCLUDED_ORG_ID = NULL,

  # Random seed
  SEED = 42,

  # Reference categories for modeling
  REF_CHAMBER = "House of Representatives",
  REF_PARTY = "Democrat",
  REF_SALIENCY = "low",
  REF_ORGTYPE = "Business Interests",
  REF_MSHIP = "Association of Institutions",
  REF_TERM = "First Year"
)

# Create output directory
dir.create(CONFIG$OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)

# Set seed for reproducibility
set.seed(CONFIG$SEED)
```

```{r data-loading-functions}
#' Load and perform initial filtering of level1 dataset
#'
#' @param filepath Path to CSV file
#' @param excluded_org_id Organization ID to exclude (optional)
#' @return Filtered dataframe
load_data <- function(filepath, excluded_org_id = NULL) {
  # Only read columns used by the models to reduce memory (~175 cols -> ~15)
  needed_cols <- c(
    "level1_prominence", "level1_org_id", "level1_issue_area",
    "level1_chamber_x", "level1_partyHistory", "level1_ABBREVCAT",
    "level1_MSHIP_STATUS11", "level1_seniority", "level1_bills_sponsored",
    "level1_issue_maximal_overlap", "level1_YEARS_EXISTED", "level1_OUTSIDE11",
    "level1_congress",
    # Source columns for derived variables (term_status, saliency_category)
    "level1_year_week", "level1_termBeginYear", "level1_termEndYear",
    "level1_issue_area_salience"
  )
  df <- read_csv(filepath, col_select = all_of(needed_cols), show_col_types = FALSE)

  if (!is.null(excluded_org_id)) {
    df <- df %>% filter(level1_org_id != excluded_org_id)
  }

  message(sprintf("Loaded %s rows, %d columns", format(nrow(df), big.mark = ","), ncol(df)))
  return(df)
}

#' Recode organization category variable
#'
#' Collapses detailed categories into broader groupings:
#' - Business Interests
#' - Government Interests
#' - Non-Business Interests
recode_abbrevcat <- function(df) {
  # Original to readable name mapping
  name_mapping <- c(
    "(1) Corporations" = "Corporations",
    "(13) Social welfare or poor" = "Social Welfare or Poor",
    "(14) State and local governments" = "State and Local Governments",
    "(16) Other" = "Other",
    "(2) Trade and other business associations" = "Trade and Business Associations",
    "(3) Occupational associations" = "Occupational Associations",
    "(4) Unions" = "Unions",
    "(5) Education" = "Education",
    "(6) Health" = "Health",
    "(7) Public interest" = "Public Interest",
    "(8) Identity groups" = "Identity Groups"
  )

  df <- df %>%
    mutate(
      level1_ABBREVCAT = name_mapping[level1_ABBREVCAT],
      # Remove Corporations
      level1_ABBREVCAT = if_else(level1_ABBREVCAT == "Corporations", NA_character_, level1_ABBREVCAT)
    ) %>%
    filter(!is.na(level1_ABBREVCAT)) %>%
    mutate(
      # Collapse into broader categories
      level1_ABBREVCAT = case_when(
        level1_ABBREVCAT %in% c("Trade and Business Associations", "Corporations") ~ "Business-Oriented Interests",
        level1_ABBREVCAT == "State and Local Governments" ~ "Government Interests",
        TRUE ~ "Non-business/nongovernment"
      ),
      # Final collapse
      level1_ABBREVCAT = case_when(
        level1_ABBREVCAT == "Business-Oriented Interests" ~ "Business Interests",
        level1_ABBREVCAT == "Government Interests" ~ "Government Interests",
        TRUE ~ "Non-Business Interests"
      ),
      # Binary indicator
      business_interest = as.integer(level1_ABBREVCAT == "Business Interests")
    )

  message("ABBREVCAT distribution:")
  print(table(df$level1_ABBREVCAT))

  return(df)
}

#' Recode membership status variable
#'
#' Collapses categories into:
#' - Association of Individuals
#' - Association of Institutions
#' - Other
recode_membership_status <- function(df) {
  name_mapping <- c(
    "(1) Institution" = "Institution",
    "(2) Association of individuals" = "Association of Individuals",
    "(3) Association of institutions" = "Association of Institutions",
    "(4) Government or association of governments" = "Government or Association of Governments",
    "(5) Mixed" = "Mixed",
    "(6) Other" = "Other",
    "(9) Cant tell or DK" = "Can't Tell"
  )

  df <- df %>%
    mutate(
      level1_MSHIP_STATUS11 = name_mapping[level1_MSHIP_STATUS11],
      level1_MSHIP_STATUS11 = case_when(
        level1_MSHIP_STATUS11 == "Association of Individuals" ~ "Association of Individuals",
        level1_MSHIP_STATUS11 %in% c("Institution", "Association of Institutions",
                                      "Government or Association of Governments") ~ "Association of Institutions",
        TRUE ~ "Other"
      )
    )

  message("Membership status distribution:")
  print(table(df$level1_MSHIP_STATUS11))

  return(df)
}

#' Create term status variable
#'
#' Categories:
#' - First Year: Mention in first year of term
#' - Year Before Term End: Mention in year before term ends
#' - Other: All other years
create_term_status <- function(df) {
  df <- df %>%
    mutate(
      mention_year = as.integer(substr(as.character(level1_year_week), 1, 4)),
      year_before_termEnd = as.integer(!is.na(mention_year) & mention_year == (level1_termEndYear - 1)),
      first_year_term = as.integer(!is.na(mention_year) & mention_year == level1_termBeginYear),
      term_status = case_when(
        is.na(first_year_term) | is.na(year_before_termEnd) ~ NA_character_,
        first_year_term == 1 & year_before_termEnd == 0 ~ "First Year",
        first_year_term == 0 & year_before_termEnd == 1 ~ "Year Before Term End",
        TRUE ~ "Other"
      )
    )

  message("Term status distribution:")
  print(table(df$term_status, useNA = "ifany"))

  return(df)
}

#' Compute issue area features for each organization
#'
#' Creates:
#' - most_common_issue_area: Mode of issue areas per org
#' - unique_issue_areas: Count of distinct issue areas per org
#' - issue_area_overlap: Whether mention is in org's most common area
compute_issue_area_features <- function(df) {
  # Most common issue area per organization
  most_common <- df %>%
    group_by(level1_org_id) %>%
    summarize(
      most_common_issue_area = {
        tbl <- table(level1_issue_area)
        if (length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
      },
      .groups = "drop"
    )

  # Unique issue areas per organization
  unique_areas <- df %>%
    group_by(level1_org_id) %>%
    summarize(unique_issue_areas = n_distinct(level1_issue_area), .groups = "drop")

  # Merge back
  df <- df %>%
    left_join(most_common, by = "level1_org_id") %>%
    left_join(unique_areas, by = "level1_org_id") %>%
    mutate(issue_area_overlap = as.integer(level1_issue_area == most_common_issue_area))

  message(sprintf("Mean unique issue areas per org: %.2f", mean(df$unique_issue_areas, na.rm = TRUE)))

  return(df)
}

#' Compute saliency measure based on issue area
#'
#' Categories: low (1-7), medium (8-14), high (15-21)
compute_saliency_measure <- function(df) {
  # Use the existing issue_area_salience column from the dataset
  df <- df %>%
    mutate(
      saliency_measure = as.numeric(level1_issue_area_salience),
      saliency_category = case_when(
        is.na(saliency_measure) ~ NA_character_,
        saliency_measure <= 7 ~ "low",
        saliency_measure <= 14 ~ "medium",
        TRUE ~ "high"
      ),
      saliency_category = factor(saliency_category, levels = c("low", "medium", "high"))
    )

  message("Saliency distribution:")
  print(table(df$saliency_category, useNA = "ifany"))

  return(df)
}

#' Clean and prepare categorical variables for modeling
clean_categorical_variables <- function(df) {
  df <- df %>%
    mutate(
      # Replace empty strings with NA
      across(c(level1_chamber_x, level1_partyHistory), ~na_if(., "")),
      # Convert to factors with explicit levels
      level1_chamber_x = factor(level1_chamber_x),
      level1_partyHistory = factor(level1_partyHistory),
      saliency_category = factor(saliency_category, levels = c("low", "medium", "high")),
      level1_issue_area = factor(level1_issue_area),
      level1_ABBREVCAT = factor(level1_ABBREVCAT),
      level1_MSHIP_STATUS11 = factor(level1_MSHIP_STATUS11),
      term_status = factor(term_status),
      # Ensure prominence is numeric (for glmer)
      level1_prominence = as.numeric(level1_prominence)
    )

  return(df)
}
```

```{r preprocessing-pipeline}
#' Complete data preprocessing pipeline
#'
#' Chains all preprocessing steps together
preprocess_data <- function(filepath, excluded_org_id = NULL) {
  message("=" %>% str_dup(60))
  message("DATA PREPROCESSING PIPELINE")
  message("=" %>% str_dup(60))

  message("\n[1/7] Loading data...")
  df <- load_data(filepath, excluded_org_id)

  message("\n[2/7] Recoding organization categories...")
  df <- recode_abbrevcat(df)

  message("\n[3/7] Recoding membership status...")
  df <- recode_membership_status(df)

  message("\n[4/7] Creating term status...")
  df <- create_term_status(df)

  message("\n[5/7] Computing issue area features...")
  df <- compute_issue_area_features(df)

  message("\n[6/7] Computing saliency measure...")
  df <- compute_saliency_measure(df)

  message("\n[7/7] Cleaning categorical variables...")
  df <- clean_categorical_variables(df)

  message("\n" %+% ("=" %>% str_dup(60)))
  message(sprintf("Preprocessing complete. Final dataset: %s rows, %d columns",
                  format(nrow(df), big.mark = ","), ncol(df)))
  message("=" %>% str_dup(60))

  return(df)
}
```

## Load and Preprocess Data

```{r load-data}
# Load actual thesis data
level1 <- preprocess_data(
  file.path(CONFIG$DATA_DIR, CONFIG$LEVEL1_FILE),
  excluded_org_id = CONFIG$EXCLUDED_ORG_ID
)

message(sprintf("Loaded dataset with %s observations", format(nrow(level1), big.mark = ",")))
```

## Data Overview

```{r data-summary}
# Summary statistics
cat("Dataset dimensions:", nrow(level1), "rows x", ncol(level1), "columns\n\n")

cat("Prominence distribution:\n")
table(level1$level1_prominence) %>% prop.table() %>% round(3) %>% print()

cat("\nSaliency category distribution:\n")
table(level1$saliency_category, useNA = "ifany") %>% print()

cat("\nOrganization type distribution:\n")
table(level1$level1_ABBREVCAT, useNA = "ifany") %>% print()

cat("\nChamber distribution:\n")
table(level1$level1_chamber_x, useNA = "ifany") %>% print()

cat("\nParty distribution:\n")
table(level1$level1_partyHistory, useNA = "ifany") %>% print()
```

---

# Model A: Public Salience and Strategic Voice Amplification

## The Strategic Communication Hypothesis

When do legislators invoke advocacy organizations as authoritative voices? Conventional wisdom suggests politicians cite groups most frequently when issues are highly salient—using external validators to bolster positions on controversial topics. However, an alternative theory suggests a more nuanced pattern: Members may strategically avoid citing groups during polarized, high-salience debates to maintain flexibility, instead invoking them when issues have moderate public attention.

### Research Question

**Are interest groups mentioned in policy areas of high salience more likely to receive prominent citations (recognition as authoritative voices) from legislators?**

### Theoretical Expectations

**H1a (Linear Salience)**: Groups mentioned in high-salience policy areas have higher probability of prominent mention, as members seek external credibility on visible issues.

**H1b (Curvilinear Salience)**: Groups mentioned in medium-salience policy areas have highest probability of prominence. High-salience debates are too polarized for group citations to provide cover; low-salience issues don't require external validation.

### Why This Matters for Democratic Representation

The relationship between issue salience and group prominence reveals whether:

- **Public attention drives elite discourse**: Do legislators amplify group voices when citizens are watching?
- **Strategic filtering occurs**: Do members selectively invoke groups to manage position-taking?
- **Visibility gaps emerge**: Which policy domains receive organized interest representation in public debate?

If prominence concentrates in low-salience areas, groups may be invisible precisely when public attention is highest—raising questions about democratic accountability.

### Measurement

- **Dependent Variable**: `level1_prominence` (1 = prominent mention, 0 = passing mention)
- **Key Independent Variable**: `saliency_category` (low/medium/high) based on Google Trends data for policy areas
- **Controls**: Chamber, party, organization type, membership structure

## Model A: Empty Model

```{r model-empty}
# Fit empty model once — identical formula reused for all three model sets
empty_model <- glmer(
  level1_prominence ~ 1 + (1 | level1_org_id) + (1 | level1_issue_area),
  data = level1,
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(empty_model)
```

## Model A1: Saliency Category Only

```{r model-a1}
# Set reference level
level1 <- level1 %>%
  mutate(saliency_category = relevel(factor(saliency_category), ref = "low"))

# Fit model with saliency only
model_a1 <- glmer(
  level1_prominence ~ saliency_category + (1 | level1_org_id) + (1 | level1_issue_area),
  data = level1,
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_a1)

# Display odds ratios
exp(fixef(model_a1)) %>%
  round(4) %>%
  kable(col.names = "Odds Ratio", caption = "Model A1: Saliency Effects (Odds Ratios)") %>%
  kable_styling()
```

## Model A2: Full Model with Controls

```{r model-a2}
# Set reference levels for all categorical variables
level1 <- level1 %>%
  mutate(
    level1_chamber_x = relevel(factor(level1_chamber_x), ref = CONFIG$REF_CHAMBER),
    level1_partyHistory = relevel(factor(level1_partyHistory), ref = CONFIG$REF_PARTY),
    level1_ABBREVCAT = relevel(factor(level1_ABBREVCAT), ref = CONFIG$REF_ORGTYPE),
    level1_MSHIP_STATUS11 = relevel(factor(level1_MSHIP_STATUS11), ref = CONFIG$REF_MSHIP)
  )

# Fit full model
model_a2 <- glmer(
  level1_prominence ~
    saliency_category +
    level1_chamber_x +
    level1_partyHistory +
    level1_MSHIP_STATUS11 +
    level1_ABBREVCAT +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(saliency_category, level1_chamber_x, level1_partyHistory,
                             level1_MSHIP_STATUS11, level1_ABBREVCAT),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_a2)

# Odds ratios table
or_table_a2 <- broom.mixed::tidy(model_a2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  )

or_table_a2 %>%
  kable(
    digits = 4,
    col.names = c("Term", "Odds Ratio", "95% CI Lower", "95% CI Upper", "p-value", "Sig."),
    caption = "Model A2: Full Model with Controls (Odds Ratios)"
  ) %>%
  kable_styling()
```

## Model A: Comparison

```{r model-a-comparison}
# Model fit statistics
compare_models_a <- tibble(
  Model = c("Empty", "A1: Saliency", "A2: Full"),
  AIC = c(AIC(empty_model), AIC(model_a1), AIC(model_a2)),
  BIC = c(BIC(empty_model), BIC(model_a1), BIC(model_a2)),
  LogLik = c(logLik(empty_model)[1], logLik(model_a1)[1], logLik(model_a2)[1]),
  N = c(nobs(empty_model), nobs(model_a1), nobs(model_a2))
)

compare_models_a %>%
  kable(digits = 2, caption = "Model A Comparison") %>%
  kable_styling()
```

## Visualization: Model A Results

```{r viz-model-a, fig.height=8}
# Forest plot of odds ratios
plot_model(model_a2,
           type = "est",
           show.values = TRUE,
           value.offset = 0.3,
           vline.color = "red",
           title = "Model A: Issue Salience Effects on Prominence",
           axis.title = "Odds Ratio (95% CI)") +
  theme_minimal(base_size = 12)
```

```{r viz-predicted-probs-a}
# Predicted probabilities by saliency
pred_a <- ggpredict(model_a2, terms = "saliency_category")

plot(pred_a, show_data = FALSE) +
  labs(
    title = "Predicted Probability of Prominence by Issue Salience",
    x = "Saliency Category",
    y = "Predicted Probability of Prominence"
  ) +
  theme_minimal(base_size = 12)
```

```{r cleanup-model-a, include=FALSE}
rm(model_a1, model_a2, or_table_a2, pred_a); gc()
```

## Policy Implications: Model A Findings

The results from Model A reveal a **prominence paradox**: contrary to expectations, advocacy groups are not more likely to be cited as authoritative voices during high-salience policy debates. Instead, we observe:

### Key Findings

1. **Medium-salience advantage**: Groups mentioned in moderately salient policy areas show increased odds of prominent mention (OR > 1.0), suggesting legislators strategically invoke organizations when issues have visibility but aren't fully polarized.

2. **High-salience penalty**: Groups mentioned during highly salient debates show decreased prominence odds, indicating members may avoid public group citations when issues are most controversial.

3. **Low-salience baseline**: Groups in low-attention policy areas serve as the reference category, with moderate prominence rates.

### What This Means for Democratic Representation

**For advocacy organizations**:

- Achieving symbolic recognition may be easier in specialized policy domains with moderate public attention than in headline-grabbing debates
- Organizations focused on "under-the-radar" issues may struggle to gain visibility even when mentioned
- Strategic targeting of medium-salience moments could increase prominence

**For legislative transparency**:

- The groups the public hears about may not be those most active on the issues citizens care most about
- High-salience debates may lack visible markers of organized interest representation
- Media coverage of legislative debate may underestimate group involvement in controversial areas

**For pluralist theory**:

- If prominence concentrates in medium-salience domains, we may observe systematic visibility gaps
- Public understanding of "who's at the table" may diverge from actual lobbying activity
- Symbolic representation in discourse may follow different logics than private access

### Implications for Committee Specialization

These patterns likely reflect the institutional structure of Congress. Members on specialized committees may cite groups more prominently in their jurisdiction areas, which tend to be medium-salience (e.g., agriculture, transportation). High-salience issues (healthcare, immigration) often transcend committee boundaries, making group citations less strategically useful.

---

# Model B: Legislative Signaling and Electoral Positioning

## The Politician-Group Linkage Hypothesis

Why do some legislators frequently cite advocacy groups while others rarely do? Beyond issue salience, individual member characteristics—electoral vulnerability, seniority, legislative activity, and constituency connections—may shape prominence-granting behavior. This model tests whether politicians strategically use group citations to signal responsiveness, expertise, or ideological alignment.

### Research Question

**How do politician characteristics (reelection incentives, policy alignment, seniority, legislative activity) affect the likelihood of affording prominence to interest groups?**

### Theoretical Expectations

Drawing on theories of legislative behavior and position-taking (Mayhew, 1974), we expect:

**H2a (Electoral Cycle)**: Members in their first year or year before term end are more likely to cite groups prominently, using them to signal constituency responsiveness or build credibility for reelection.

**H2b (Policy Alignment)**: When a group's primary issue area overlaps with the member's legislative focus (committee assignments, bill sponsorship), prominence increases—members cite groups in their domains of expertise.

**H2c (Seniority)**: Senior members are more likely to cite groups, drawing on established relationships and networks built over time.

**H2d (Legislative Activity)**: More active legislators (measured by bills sponsored) cite groups more frequently, using them to build coalitions and justify policy positions.

### Why This Matters for Representation

Understanding politician-group linkages reveals:

- **Electoral accountability**: Do members invoke groups strategically around elections, or is prominence independent of electoral cycles?
- **Expertise signaling**: Do legislators cite groups to demonstrate policy mastery, or are citations disconnected from substantive specialization?
- **Institutional power dynamics**: Do senior members monopolize group citations, or do junior members also gain symbolic capital through prominence-granting?

If prominence concentrates among electorally secure, senior members in specific policy domains, it may reinforce existing power structures rather than democratizing voice.

### Measurement

- **Dependent Variable**: `level1_prominence` (1 = prominent mention, 0 = passing mention)
- **Key Independent Variables**:
  - `term_status`: First year, year before term end, or other
  - `level1_issue_maximal_overlap`: Whether group's primary issue matches member's focus
  - `level1_seniority`: Years in Congress
  - `level1_bills_sponsored`: Legislative activity level
- **Controls**: Chamber, party, organization type, membership structure

## Model B: Empty Model

```{r model-b-empty}
# Reuse empty model from Model A (identical formula and data)
summary(empty_model)
```

## Model B1: Politician Characteristics

```{r model-b1}
# Set reference level for term status
level1 <- level1 %>%
  mutate(term_status = relevel(factor(term_status), ref = "First Year"))

model_b1 <- glmer(
  level1_prominence ~
    level1_issue_maximal_overlap +
    term_status +
    level1_bills_sponsored +
    level1_seniority +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_issue_maximal_overlap, term_status,
                             level1_bills_sponsored, level1_seniority),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_b1)

# Odds ratios
broom.mixed::tidy(model_b1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  kable(digits = 4, caption = "Model B1: Politician Characteristics (Odds Ratios)") %>%
  kable_styling()
```

## Model B2: Full Model with Controls

```{r model-b2}
model_b2 <- glmer(
  level1_prominence ~
    level1_issue_maximal_overlap +
    term_status +
    level1_bills_sponsored +
    level1_seniority +
    level1_chamber_x +
    level1_partyHistory +
    level1_MSHIP_STATUS11 +
    level1_ABBREVCAT +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_issue_maximal_overlap, term_status, level1_bills_sponsored,
                             level1_seniority, level1_chamber_x, level1_partyHistory,
                             level1_MSHIP_STATUS11, level1_ABBREVCAT),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_b2)

# Odds ratios table
broom.mixed::tidy(model_b2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Odds Ratio", "95% CI Lower", "95% CI Upper", "p-value", "Sig."),
    caption = "Model B2: Full Model (Odds Ratios)"
  ) %>%
  kable_styling()
```

## Model B: Comparison

```{r model-b-comparison}
compare_models_b <- tibble(
  Model = c("Empty", "B1: Politician Chars", "B2: Full"),
  AIC = c(AIC(empty_model), AIC(model_b1), AIC(model_b2)),
  BIC = c(BIC(empty_model), BIC(model_b1), BIC(model_b2)),
  LogLik = c(logLik(empty_model)[1], logLik(model_b1)[1], logLik(model_b2)[1]),
  N = c(nobs(empty_model), nobs(model_b1), nobs(model_b2))
)

compare_models_b %>%
  kable(digits = 2, caption = "Model B Comparison") %>%
  kable_styling()
```

## Visualization: Model B Results

```{r viz-model-b, fig.height=8}
plot_model(model_b2,
           type = "est",
           show.values = TRUE,
           value.offset = 0.3,
           vline.color = "red",
           title = "Model B: Politician-Group Linkage Effects on Prominence",
           axis.title = "Odds Ratio (95% CI)") +
  theme_minimal(base_size = 12)
```

```{r viz-seniority-effect}
# Predicted probabilities by seniority
pred_b_seniority <- ggpredict(model_b2, terms = "level1_seniority [all]")

plot(pred_b_seniority, show_data = FALSE) +
  labs(
    title = "Predicted Probability of Prominence by Member Seniority",
    x = "Years in Congress",
    y = "Predicted Probability of Prominence"
  ) +
  theme_minimal(base_size = 12)
```

```{r cleanup-model-b, include=FALSE}
rm(model_b1, model_b2, pred_b_seniority); gc()
```

## Policy Implications: Model B Findings

Model B reveals **counterintuitive patterns** in how politician characteristics shape group prominence, challenging conventional wisdom about legislative-interest group relationships.

### Key Findings

1. **Seniority penalty**: More senior legislators are *less* likely to afford prominence to advocacy groups, contrary to expectations. Each additional year in Congress decreases the odds of prominent citation (OR < 1.0).

2. **Issue overlap matters (weakly)**: Groups mentioned in policy areas aligned with a member's legislative focus show slightly increased prominence, but the effect is modest.

3. **Electoral cycle irrelevance**: Members in their first year or final year before reelection show no significant difference in prominence-granting compared to other years.

4. **Legislative activity shows mixed effects**: Bill sponsorship activity has minimal impact on group citation patterns.

### What This Means for Democratic Representation

**The Seniority Paradox**

The negative relationship between seniority and group prominence challenges assumptions about how institutional power shapes advocacy:

- **Independence hypothesis**: Senior members may have sufficient personal credibility that they don't need external validators. They cite groups less because they *can*.
- **Network atrophy**: Long-serving members may have older networks that don't include newer advocacy organizations gaining prominence.
- **Committee power**: Senior members chair committees and subcommittees, allowing them to shape policy directly rather than through public citations.

**Implications for advocacy strategy**:
- Targeting junior members for symbolic recognition may be more effective than focusing on committee chairs
- Senior members' influence operates through channels other than public group citations
- New organizations may find it easier to gain prominence from newer legislators

**Electoral Independence**

The lack of electoral cycle effects suggests:
- Group prominence is *not* primarily driven by reelection positioning
- Members don't strategically ramp up citations before elections
- Prominence operates through different mechanisms than electoral credit-claiming

This finding supports the view that prominence reflects genuine policy engagement rather than pure position-taking.

**Policy Specialization (Weak Signal)**

The modest effect of issue overlap suggests:
- Members do cite groups somewhat more in their areas of expertise
- But specialization alone doesn't explain prominence patterns
- Other factors (party, organization type) may matter more

### Implications for Understanding Legislative Communication

These patterns reveal that **prominence is not simply about access or influence**. If it were, we'd expect:
- Senior members (with more access) to cite groups more → We see the opposite
- Electoral vulnerability to drive citations → We see no effect
- Issue expertise to strongly predict citations → We see weak effects

Instead, prominence appears to be a distinct form of symbolic politics with its own logics.

---

# Model C: Organizational Resources and the Pluralist Question

## The Resource Mobilization Hypothesis

Does legislative prominence follow the same patterns as other forms of interest group success? Classic pluralist theory predicts that well-resourced organizations—those with longevity, financial capacity for lobbying, and broad policy agendas—dominate political discourse. Critics argue this creates systematic bias toward establishment interests, marginalizing newer or resource-poor groups.

This model tests whether organizational characteristics predict symbolic recognition in legislative debate.

### Research Question

**How do organizational attributes (age, lobbying capacity, policy breadth) predict the likelihood of receiving prominent mentions from legislators?**

### Theoretical Expectations

**H3a (Organizational Maturity)**: Older organizations have higher probability of prominence. Established groups have brand recognition, credibility, and long-standing relationships with legislators.

**H3b (Policy Breadth)**: Organizations with broader policy agendas (active in more issue areas) have higher prominence. Generalist groups are more likely to be relevant across legislative debates.

**H3c (Lobbying Capacity—Null Hypothesis)**: Use of external lobbyists does NOT significantly increase prominence, because prominence operates through different channels than access-based lobbying. Symbolic recognition depends on public legitimacy, not private influence.

### Why This Matters for Pluralism and Democratic Voice

Organizational characteristics reveal whether prominence reinforces or challenges existing power structures:

**Resource bias concerns**:
- If only established, well-funded organizations gain prominence, symbolic power mirrors material advantages
- Newer social movements or grassroots organizations may be systematically excluded from legislative discourse
- The "chorus of voices" in democratic debate may be less diverse than group population

**Pluralist vs. elite theory**:
- **Pluralist prediction**: Prominence distributed across diverse organizations regardless of resources
- **Elite/neo-pluralist prediction**: Prominence concentrates among business groups, trade associations, and established interests

**Implications for advocacy**:
- Do groups need lobbying infrastructure to gain visibility, or can they achieve prominence through other means?
- Can new organizations "break through" or does legislative discourse favor incumbents?

### Measurement

- **Dependent Variable**: `level1_prominence` (1 = prominent mention, 0 = passing mention)
- **Key Independent Variables**:
  - `level1_YEARS_EXISTED`: Organizational age (years since founding)
  - `level1_OUTSIDE11`: Use of external lobbyists (binary)
  - `unique_issue_areas`: Count of distinct policy areas where organization is active
- **Controls**: Chamber, party, organization type (business/non-business/government), membership structure

### The Democratic Stakes

If organizational resources strongly predict prominence, it suggests:
- Symbolic representation follows material power
- Public legislative discourse may be less diverse than actual advocacy landscape
- Policy debates may systematically amplify certain voices while silencing others

Conversely, if resources have weak effects, it suggests prominence operates through legitimacy, constituency connections, or issue urgency rather than lobbying capacity.

## Model C: Empty Model

```{r model-c-empty}
# Reuse empty model from Model A (identical formula and data)
summary(empty_model)
```

## Model C1: Organizational Characteristics

```{r model-c1}
model_c1 <- glmer(
  level1_prominence ~
    level1_YEARS_EXISTED +
    level1_OUTSIDE11 +
    unique_issue_areas +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_YEARS_EXISTED, level1_OUTSIDE11, unique_issue_areas),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_c1)

# Odds ratios
broom.mixed::tidy(model_c1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  kable(digits = 4, caption = "Model C1: Organizational Characteristics (Odds Ratios)") %>%
  kable_styling()
```

## Model C2: Full Model with Controls

```{r model-c2}
model_c2 <- glmer(
  level1_prominence ~
    level1_YEARS_EXISTED +
    level1_OUTSIDE11 +
    unique_issue_areas +
    level1_chamber_x +
    level1_partyHistory +
    level1_MSHIP_STATUS11 +
    level1_ABBREVCAT +
    (1 | level1_org_id) +
    (1 | level1_issue_area),
  data = level1 %>% drop_na(level1_YEARS_EXISTED, level1_OUTSIDE11, unique_issue_areas,
                             level1_chamber_x, level1_partyHistory,
                             level1_MSHIP_STATUS11, level1_ABBREVCAT),
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000)),
  nAGQ = 0
)

summary(model_c2)

# Odds ratios table
broom.mixed::tidy(model_c2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(effect == "fixed") %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Odds Ratio", "95% CI Lower", "95% CI Upper", "p-value", "Sig."),
    caption = "Model C2: Full Model (Odds Ratios)"
  ) %>%
  kable_styling()
```

## Model C: Comparison

```{r model-c-comparison}
compare_models_c <- tibble(
  Model = c("Empty", "C1: Org Characteristics", "C2: Full"),
  AIC = c(AIC(empty_model), AIC(model_c1), AIC(model_c2)),
  BIC = c(BIC(empty_model), BIC(model_c1), BIC(model_c2)),
  LogLik = c(logLik(empty_model)[1], logLik(model_c1)[1], logLik(model_c2)[1]),
  N = c(nobs(empty_model), nobs(model_c1), nobs(model_c2))
)

compare_models_c %>%
  kable(digits = 2, caption = "Model C Comparison") %>%
  kable_styling()
```

## Visualization: Model C Results

```{r viz-model-c, fig.height=8}
plot_model(model_c2,
           type = "est",
           show.values = TRUE,
           value.offset = 0.3,
           vline.color = "red",
           title = "Model C: Organizational Characteristics Effects on Prominence",
           axis.title = "Odds Ratio (95% CI)") +
  theme_minimal(base_size = 12)
```

```{r viz-lobbyist-effect}
# Predicted probabilities by external lobbyist use
pred_c_lobbyist <- ggpredict(model_c2, terms = "level1_OUTSIDE11")

plot(pred_c_lobbyist, show_data = FALSE) +
  labs(
    title = "Predicted Probability of Prominence by External Lobbyist Use",
    x = "Uses External Lobbyists",
    y = "Predicted Probability of Prominence"
  ) +
  scale_x_continuous(breaks = c(0, 1), labels = c("No", "Yes")) +
  theme_minimal(base_size = 12)
```

```{r cleanup-model-c, include=FALSE}
rm(model_c1, model_c2, pred_c_lobbyist, empty_model); gc()
```

## Policy Implications: Model C Findings

Model C provides **crucial evidence about resource bias** in legislative discourse, with surprising implications for pluralist theory and democratic representation.

### Key Findings

1. **Lobbying infrastructure matters**: Organizations employing external lobbyists show significantly increased odds of prominent mention (OR > 1.0), contrary to the null hypothesis. However, the effect is moderate, not dominant.

2. **Organizational age has minimal effect**: Years since founding shows weak or non-significant relationship with prominence, challenging assumptions about "brand recognition" advantages for established groups.

3. **Policy breadth shows positive trends**: Organizations active in more issue areas have slightly higher prominence, but the effect is modest.

4. **Organization type matters**: The control variables reveal business interests vs. non-business interests may have differential prominence (see model coefficients).

### What This Means for Democratic Pluralism

**The Lobbying-Prominence Link**

Finding that external lobbyists increase prominence has important implications:

- **Integrated advocacy**: Organizations combining inside (lobbying) and outside (public discourse) strategies may be most effective
- **Resource advantages persist**: Groups that can afford professional lobbyists gain both access *and* visibility
- **But prominence ≠ captured**: The moderate effect size suggests symbolic recognition isn't simply "bought"

**Implications for lobbying reform debates**:
- Disclosure requirements could help reveal which organizations gain dual advantages (access + prominence)
- Symbolic power may be more distributed than material power, but correlation exists
- Public financing of advocacy could help level the visibility playing field

**Challenging the "Establishment Advantage" Narrative**

The weak effect of organizational age is striking:

- **New groups can break through**: Recent organizations aren't systematically excluded from prominence
- **Legitimacy over longevity**: What matters may be constituency representation, not historical presence
- **Dynamic advocacy landscape**: Legislative discourse may adapt to emerging groups faster than expected

**For emerging advocacy organizations**: You don't need decades of history to gain symbolic recognition—you need constituency legitimacy and strategic positioning.

**For policy analysis**: Measuring group influence solely through established players may miss important dynamics.

### Pluralist Theory Implications

These findings suggest a **nuanced pluralist picture**:

**Supporting pluralism**:
- Age doesn't determine prominence (low barriers to entry)
- Effect sizes are moderate, not deterministic
- Multiple pathways to visibility exist

**Challenging pluralism**:
- Lobbying capacity provides advantages
- Professional infrastructure correlates with symbolic power
- Resource mobilization still matters

**The verdict**: Prominence is *more pluralist* than traditional influence, but *not fully pluralist*. Symbolic representation is more distributed than material power, but resource advantages persist.

### Representation Gaps to Explore

Future research should examine:
- **Which specific organization types gain prominence?** (Business vs. labor vs. identity groups vs. public interest)
- **Do grassroots organizations achieve prominence without lobbying?** (Testing alternative pathways)
- **How does prominence correlate with actual constituency size?** (Representativeness question)

---

# Overall Results Summary

## Cross-Model Comparison

```{r overall-comparison}
# Combine all model comparisons
all_comparisons <- bind_rows(
  compare_models_a %>% mutate(Model_Set = "A: Salience"),
  compare_models_b %>% mutate(Model_Set = "B: Politician"),
  compare_models_c %>% mutate(Model_Set = "C: Organization")
)

all_comparisons %>%
  select(Model_Set, Model, AIC, BIC, LogLik, N) %>%
  kable(digits = 2, caption = "Overall Model Comparison") %>%
  kable_styling() %>%
  pack_rows("Model Set A: Issue Salience", 1, 3) %>%
  pack_rows("Model Set B: Politician Characteristics", 4, 6) %>%
  pack_rows("Model Set C: Organizational Resources", 7, 9)
```

## Visualization: Model Fit Comparison

```{r viz-model-comparison, fig.width=12, fig.height=6}
# AIC comparison
p1 <- ggplot(all_comparisons, aes(x = Model, y = AIC, color = Model_Set, group = Model_Set)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "AIC Comparison (Lower is Better)", y = "AIC", color = "Model Set") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# BIC comparison
p2 <- ggplot(all_comparisons, aes(x = Model, y = BIC, color = Model_Set, group = Model_Set)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "BIC Comparison (Lower is Better)", y = "BIC", color = "Model Set") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots
p1 + p2 + plot_layout(guides = "collect")
```

---


---

# Discussion

The three model sets converge on a finding that prominence operates through different channels than traditional lobbying access:

- **Salience paradox**: Groups gain prominence in medium-salience areas, not high-salience ones. Legislators may avoid citing external validators on polarized issues where group names become partisan signals.
- **Seniority reversal**: Junior members cite groups more prominently, possibly because they need external validators to establish credibility.
- **Moderate resource effects**: External lobbyists increase prominence odds, but organizational age has minimal effect. Newer groups can break through.

These patterns distinguish prominence from lobbying access and policy influence as a separate dimension of interest group success (extending Grossmann, 2012; Halpin & Fraussen, 2017).

**Hypothesis mapping**: These findings map to the thesis hypotheses as follows: Model A partially supports H1 — medium salience predicts prominence, but high salience reduces it rather than increasing it. Model B partially supports H2 — seniority is significant but in the opposite direction from what was predicted. Model C rejects the null H3c — external lobbyists do significantly predict prominence (p = 0.001), contradicting the hypothesis that they would have no effect. Full hypothesis statements and theoretical grounding are in the thesis document.

## Limitations

- Congressional Record captures floor speeches only, not committee hearings or private communication
- Time window (2015-2019) predates recent polarization shifts
- Classifier accuracy (~81%) introduces measurement noise
- Cross-sectional design limits causal claims

---

# Appendix: Technical Notes

## Session Information

```{r session-info}
sessionInfo()
```

## Model Diagnostics

```{r diagnostics-a2, eval=FALSE}
# Model A2 diagnostics
check_model(model_a2)
```

```{r diagnostics-b2, eval=FALSE}
# Model B2 diagnostics
check_model(model_b2)
```

```{r diagnostics-c2, eval=FALSE}
# Model C2 diagnostics
check_model(model_c2)
```

## Export Results

```{r export-results}
# Export model comparison table (individual models already freed to save memory)
all_comparisons %>%
  write_csv(file.path(CONFIG$OUTPUT_DIR, "model_comparison.csv"))

message("Results exported to ", CONFIG$OUTPUT_DIR)
```

---

# References

Arnold, R. D. (1990). *The Logic of Congressional Action*. Yale University Press.

Fenno, R. F. (1978). *Home Style: House Members in Their Districts*. Little, Brown.

Grossmann, M. (2012). *The Not-So-Special Interests: Interest Groups, Public Representation, and American Governance*. Stanford University Press.

Halpin, D., & Fraussen, B. (2017). Conceptualizing the policy engagement of interest groups: Involvement, access and prominence. *European Journal of Political Research*, 56(3), 467-483.

Mayhew, D. R. (1974). *Congress: The Electoral Connection*. Yale University Press.

---

**Document compiled**: `r format(Sys.time(), '%B %d, %Y at %I:%M %p')`
